Can Auxiliary Indicators Improve COVID-19 Forecasting and Hotspot Prediction?

### 2021-10-05 forecast/paper/paper.tex
Pass over paper
- Mostly just small copyedits
- Moved reference to the supplement's new statistical significance
  tests the resuls section

While forecasting carries a rich literature that offers a wide range of
techniques, see e.g., \cite{Hyndman:2018}, we purposely constrain ourselves to
very simple models, avoiding common enhancements such as order selection,
correction of outliers/anomalies in the data, and inclusion of regularization or
nonlinearities.
% Addition in response to Reviewer 1
For the same reason, we do not account for any environmental factors that may
well aid in forecasting---age structure, variants, holidays,
reporting patterns, vaccination, public health measures, etc.
% End addition
That said, analyses of forecasts submitted to the COVID-19
Forecast Hub \cite{ForecastHub} by a large community of modelers have shown that
simple, robust models have consistently been among the best-performing over the
pandemic, including time series models similar to those we
consider in what follows.
% Added in response to Reviewer 2
\cite{Cramer:2021} investigates the performance of submitted forecasts for
state-level deaths and shows that time series models perform among the top 5.
Our internal evaluations suggest that the same is true (at least, if not to a
greater extent) for case forecasting.

@@

Finally, it is worth noting that examining the importance of additional features
for prediction is a core question in inferential statistics and econometrics,
with work dating back to at least \cite{Granger:1969}. Still today, drawing
rigorous inference based on predictions, without (or with lean) assumptions, is
an active field of research from both the applied and theoretical angles;
see, e.g., \cite{Diebold:2002, McCraken:2007, Diebold:2015, Stokes:2017,
  Lei:2018, Rinaldo:2019, Williamsom:2020, Zhang:2020, Dai:2021, Fryer:2021}.
Our take on this problem is in line with much of this literature; however, in
order to avoid making any explicit assumptions, we do not attempt to make formal 
significance statements, and instead, broadly examine the stability of our
conclusions with respect to numerous modes of analysis.
% Addition upon review
In the Supplement we examine
two assumption-lean
methods for assessing the statistical significance of our main results.

@@

\paragraph{Dynamic Re-Training}

@@

In our analysis, we let the prediction date $t$ run over each day in between
early/mid June and December 31, 2020.  The precise start date differs for 
forecasting and hotspots prediction; for each task it was chosen to be the
earliest date at which the data needed to train all models was available, which
ends up being (per our setup, with 21 days of training data and lagged values 
of signals for features, as we will detail shortly) June 9, 2020 for
forecasting, and June 16, 2020 for hotspot prediction. (The bottleneck here is
the CTIS-CLI-in-community signal, which does not exist before early April 2020,
when the survey was first launched). The end date was chosen again with a 
consideration to align both tasks as best as possible,
% added to emphasize 
and to coincide with the
introduction of vaccinations.
% Removed: this is no longer true unfortunately.
% because few hotspots
% exist post December 31, 2020, due to the general and gradual decline of the
% pandemic in 2021.

@@

For forecasting, we evaluate the probabilistic forecasts produced by the
quantile models in \eqref{eq:forecast_ar} and \eqref{eq:forecast_ar_x} using 
\textit{weighted interval score} (WIS), a quantile-based scoring rule; see e.g.,
\cite{Gneiting:2007}.  WIS is a proper score, which means that its expectation 
is minimized by the population quantiles of the target variable.  The use of WIS
in COVID-19 forecast scoring is discussed in \cite{Bracher:2021}; WIS is also
the main evaluation metric used in the COVID-19 Forecast Hub. More broadly, the
specific form of WIS used here is a standard metric in the forecasting community
for evaluating quantile-based probabilistic forecasts, just as mean squared
forecast error is standard for point forecasts.

@@

Here, and in what follows, we will use ``AR'' to refer to the pure 
autoregressive model both in forecasting, \eqref{eq:forecast_ar}, and in hotspot
prediction, \eqref{eq:hotspot_ar} (the reference to the prediction task should
always be clear from the context). We will also use the name of an auxiliary 
indicator---namely ``CHNG-CLI'',  ``CHNG-COVID'', ``CTIS-CLI-in-community'',
``DV-CLI'', or ``Google-AA''---interchangeably with the model in forecasting, 
\eqref{eq:forecast_ar_x}, or hotspot prediction, \eqref{eq:hotspot_ar_x}, that
uses this particular indicator as a feature (the meaning should be clear from
the context).  So, for example, the CHNG-CLI model in forecasting is the one in  
\eqref{eq:forecast_ar_x} that sets $X_{\ell,t}$ to be the value of the CHNG-CLI 
indicator at location $\ell$ and time $t$.
% fits the coefficients based on the training data, and makes its predictions by
% setting $X_{\ell,t}$ to the appropriate CHNG-CLI values.  
Finally, we use the term ``indicator model'' to refer to any one of the ten
models of the form \eqref{eq:forecast_ar_x} or \eqref{eq:hotspot_ar_x} (five
from each of the forecasting and hotspot prediction tasks).

@@

\item In a more detailed analysis of forecasting performance, we find that the
  indicator models tend to be better than AR when case rates are flat or
  decreasing (most notable in CHNG-COVID and CTIS-CLI-in-community), but they
  are worse than AR when case rates are increasing (this is most notable in
  CHNG-CLI and DV-CLI). More rarely does an indicator model tend to beat AR when
  case rates are increasing, but there appears to be some evidence of this for
  the Google-AA model. 
  % In an increasing period, when an indicator model is worse than AR, it tends
  % to underpredict relative to AR (its median forecast is lower).  In a flat
  % or decreasing period, when an indicator model is better than AR, AR tends
  % to overpredict relative to it (the median AR forecast is higher).

@@

Finally, to quantify the importance of training and making predictions using
proper vintage data, we ran a parallel set of forecasting and hotspot prediction 
experiments using finalized data. The results, given in the supplement, show
that training and making predictions on finalized data can result in overly
optimistic estimates of true test-time performance (up to 10\% better in terms
of average WIS or AUC). Furthermore, since indicators can have greatly different
backfill profiles, the use of finalized data in retrospective evaluations
changes the relative ranking of models.  For example, CHNG-CLI and DV-CLI,
when trained on finalized data, perform very similarly in forecasting.
This makes sense since they are both claims-based indicators that are supposedly
measuring the same thing.  However, DV-CLI outperforms CHNG-CLI on vintage
data, reflecting its has a less severe backfill profile.

@@

\begin{figure*}[t]
  \includegraphics[width=\textwidth]{fig/fcast-hot-combo-1.pdf}
  \caption{Main results for both tasks. Left: average WIS for each forecast
    model, over all forecast dates and all HRRs, divided by the average WIS
    achieved by a baseline model (a probabilistic version of the flat-line
    forecaster).  Right: area under the curve for each hotspot prediction model,
    calculated over all prediction dates and all HRRs.  Here and in all figures
    we abbreviate CTIS-CLI-in-community by CTIS-CLIIC.  
  }  
  \label{fig:forecast}
\end{figure*}

@@

We can also see from the figure that CHNG-COVID and DV-CLI offer the biggest
gains over AR at small ahead values, followed by CTIS-CLI-in-community and
Google-AA, with the former providing the biggest gains at large ahead values.
The CHNG-CLI model performs basically the same as AR.  This is likely due to the
fact that CHNG-CLI suffers from volatility due to backfill.  The evidence for
this explanation is twofold: (1) the CHNG-CLI model benefits from a more robust
method of aggregating WIS (geometric mean; shown in the supplement); and (2)
when we train and make predictions on finalized data, it handily beats AR, on
par with the best-performing models (also shown in the supplement).

Figure~\ref{fig:forecast} (right panel) displays the results for hotspot
prediction, again stratified by ahead value and averaged over all HRRs and
prediction dates.  We can see many similarities to the forecasting results (now
larger AUC is better). For example, CHNG-COVID and DV-CLI offer the biggest
improvement over AR, and all models, including AR, degrade in performance
towards the baseline (in this context, a classifier based on random guessing,
which achieves an AUC of 0.5) as the ahead values grow, illustrating the
difficulty of the hotspot prediction problem. A clear difference, however, is
that the CHNG-CLI model performs quite well in hotspot prediction, close to the
best-performing indicator models for many of the ahead values. This may be
because volatility in the CHNG-CLI indicator plays less of a role in the
associated logistic model's predicted probabilities (in general, a sigmoid
function can absorb a lot of the variability in its input).

@@

To test this hypothesis, we reran all of the prediction experiments but with
$X_{\ell,t}$ in the each indicator models replaced with suitable random noise
(bootstrap samples from a signal's history).  The results, shown and explained
more precisely in the supplement, are vastly different (worse) than the original
set of results.  In both forecasting and hotspot prediction, the ``fake''
indicator models offered essentially no improvement over the pure AR model,
which strongly rejects (informally speaking) the implicit regularization
hypothesis.

@@

The supplement pursues this analysis further.  For example, we examine
classification accuracy and log-likelihood for the hotspot task and find a
similar phenomenon: the indicators considerably improve accuracy or
log-likelihood during flat or down periods, with more mixed behavior during up
periods when CHNG-CLI, CHNG-COVID, and DV-CLI, in particular, lead to decreased  
performance.

>>>>

While forecasting carries a rich literature that offers a wide range of
techniques, see, e.g., \cite{Hyndman:2018}, we purposely constrain ourselves to 
very simple models, avoiding common enhancements such as order selection,
correction of outliers/anomalies in the data, and inclusion of regularization or
nonlinearities. For the same reason, we do not account for other factors that
may well aid in forecasting, such as age-specific effects, holiday adjustments,
public health mandates, and so on. That said, analyses of forecasts submitted
to the COVID-19 Forecast Hub \cite{ForecastHub} by a large community of modelers 
have shown that simple, robust models have consistently been among the
best-performing over the pandemic, including time series models similar the ones
we use in this paper. See, e.g., \cite{Cramer:2021}, for a detailed study
of death forecasting. Our internal evaluations suggest that the same is true (at
least, if not to a greater extent) for case forecasting.

@@

Finally, it is worth noting that examining the importance of additional features
for prediction is a core question in inferential statistics and econometrics,
with work dating back to at least \cite{Granger:1969}. Still today, drawing
rigorous inference based on predictions, without (or with lean) assumptions, is
an active field of research from both the applied and theoretical angles;
see, e.g., \cite{Diebold:2002, McCraken:2007, Diebold:2015, Stokes:2017,
  Lei:2018, Rinaldo:2019, Williamsom:2020, Zhang:2020, Dai:2021, Fryer:2021}.
Our take on this problem is in line with much of this literature; however, in
order to avoid making any explicit assumptions, we do not attempt to make formal 
significance statements, and instead, broadly examine the stability of our
conclusions with respect to numerous modes of analysis.

@@

\paragraph{Dynamic Retraining}

@@

In our analysis, we let the prediction date $t$ run over each day in between
early/mid June and December 31, 2020.  The precise start date differs for 
forecasting and hotspots prediction; for each task it was chosen to be the
earliest date at which the data needed to train all models was available, which
ends up being (per our setup, with 21 days of training data and lagged values 
of signals for features, as we will detail shortly) June 9, 2020 for
forecasting, and June 16, 2020 for hotspot prediction. (The bottleneck here is
the CTIS-CLI-in-community signal, which does not exist before early April 2020,
when the survey was first launched). The end date was chosen again with a 
consideration to align both tasks as best as possible, and because very few
hotspots exist for the several few months of 2021, due to the general and
gradual decline of the pandemic during this period.

@@

For forecasting, we evaluate the probabilistic forecasts produced by the
quantile models in \eqref{eq:forecast_ar} and \eqref{eq:forecast_ar_x} using
\textit{weighted interval score} (WIS), a quantile-based scoring rule; see,
e.g., \cite{Gneiting:2007}.  WIS is a proper score, which means that its
expectation is minimized by the population quantiles of the target variable.
The use of WIS in COVID-19 forecast scoring is discussed in \cite{Bracher:2021};
WIS is also the main evaluation metric used in the COVID-19 Forecast Hub. More
broadly, the specific form of WIS used here is a standard metric in the
forecasting community for evaluating quantile-based probabilistic forecasts,
just as mean squared forecast error is standard for point forecasts.

@@

Here, and in what follows, we will use ``AR'' to refer to the pure 
autoregressive model both in forecasting, \eqref{eq:forecast_ar}, and in hotspot
prediction, \eqref{eq:hotspot_ar} (the reference to the prediction task should
always be clear from the context). We will also use the name of an auxiliary 
indicator---namely ``CHNG-CLI'',  ``CHNG-COVID'', ``CTIS-CLI-in-community'',
``DV-CLI'', or ``Google-AA''---interchangeably with the model in forecasting, 
\eqref{eq:forecast_ar_x}, or hotspot prediction, \eqref{eq:hotspot_ar_x}, that
uses this particular indicator as a feature (the meaning should be clear from
the context).  So, for example, the CHNG-CLI model in forecasting is the one in  
\eqref{eq:forecast_ar_x} that sets $X_{\ell,t}$ to be the value of the CHNG-CLI
indicator at location $\ell$ and time $t$.  Finally, we use the term ``indicator
model'' to refer to any one of the ten models of the form
\eqref{eq:forecast_ar_x} or \eqref{eq:hotspot_ar_x} (five from each of the
forecasting and hotspot prediction tasks).

@@

\item In a more detailed analysis of forecasting performance, we find that the
  indicator models tend to be better than AR when case rates are flat or
  decreasing (most notable in CHNG-COVID and CTIS-CLI-in-community), but can be
  worse than AR when case rates are increasing (this is most notable in CHNG-CLI
  and DV-CLI). More rarely does an indicator model tend to beat AR when case
  rates are increasing, but there appears to be some evidence of this for the
  Google-AA model.  
  % In an increasing period, when an indicator model is worse than AR, it tends
  % to underpredict relative to AR (its median forecast is lower).  In a flat
  % or decreasing period, when an indicator model is better than AR, AR tends
  % to overpredict relative to it (the median AR forecast is higher).

@@

Finally, to quantify the importance of training and making predictions using
proper vintage data, we ran a parallel set of forecasting and hotspot prediction
experiments using finalized data. The results, given in the Supplementary
Materials, show that training and making predictions on finalized data can
result in overly optimistic estimates of true test-time performance (up to 10\%
better in terms of average WIS or AUC). Furthermore, since indicators can have
greatly different backfill profiles, the use of finalized data in retrospective
evaluations changes the relative ranking of models.  For example, CHNG-CLI and
DV-CLI, when trained on finalized data, perform very similarly in forecasting.
This makes sense since they are both claims-based indicators that are supposedly
measuring the same thing.  However, DV-CLI outperforms CHNG-CLI on vintage data,
reflecting its has a less severe backfill profile.

The Supplement provides a number of other additional analyses; for example, we
examine two assumption-lean methods for assessing the statistical significance
of our main results.

@@

\begin{figure*}[t]
  \centering
  \includegraphics[width=0.95\textwidth]{fig/fcast-hot-combo-1.pdf}
  \caption{Main results for both tasks. Left: average WIS for each forecast
    model, over all forecast dates and all HRRs, divided by the average WIS
    achieved by a baseline model (a probabilistic version of the flat-line
    forecaster).  Right: area under the curve for each hotspot prediction model,
    calculated over all prediction dates and all HRRs.  Here and in all figures
    we abbreviate CTIS-CLI-in-community by CTIS-CLIIC.  
  }  
  \label{fig:forecast}
\end{figure*}

@@

We can also see from the figure that CHNG-COVID and DV-CLI offer the biggest
gains over AR at small ahead values, followed by CTIS-CLI-in-community and
Google-AA, with the former providing the biggest gains at large ahead values.
The CHNG-CLI model performs basically the same as AR.  This is likely due to the
fact that CHNG-CLI suffers from volatility due to backfill.  The evidence for
this explanation is twofold: (1) the CHNG-CLI model benefits from a more robust
method of aggregating WIS (geometric mean; shown in the Supplementary
Materials); and (2) when we train and make predictions on finalized data, it
handily beats AR, on par with the best-performing models (also shown in the
Supplement).

Figure~\ref{fig:forecast} (right panel) displays the results for hotspot
prediction, again stratified by ahead value and averaged over all HRRs and
prediction dates.  We can see many similarities to the forecasting results
(note: now larger AUC is better). For example, CHNG-COVID and DV-CLI offer the
biggest improvement over AR, and all models, including AR, degrade in
performance towards the baseline (in this context, a classifier based on random
guessing, which achieves an AUC of 0.5) as the ahead values grow, illustrating
the difficulty of the hotspot prediction problem. A clear difference, however,
is that the CHNG-CLI model performs quite well in hotspot prediction, close to
the best-performing indicator models for many of the ahead values. This may be
because volatility in the CHNG-CLI indicator plays less of a role in the
associated logistic model's predicted probabilities (in general, a sigmoid
function can absorb a lot of the variability in its input).

@@

To test this hypothesis, we reran all of the prediction experiments but with
$X_{\ell,t}$ in the each indicator models replaced with suitable random noise
(bootstrap samples from a signal's history).  The results, shown and explained
more precisely in the Supplementary Materials, are vastly different (worse) than
the original set of results.  In both forecasting and hotspot prediction, the
``fake'' indicator models offered essentially no improvement over the pure AR
model, which strongly rejects (informally speaking) the implicit regularization
hypothesis.

@@

The Supplement pursues this analysis further.  For example, we examine
classification accuracy and log-likelihood for the hotspot task and find a
similar phenomenon: the indicators considerably improve accuracy or
log-likelihood during flat or down periods, with more mixed behavior during up
periods when CHNG-CLI, CHNG-COVID, and DV-CLI, in particular, lead to decreased  
performance.
